---
title: "CKMRsim"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This package implements Monte Carlo methods
(including efficient importance sampling approaches) for assessing the false positive and false
negative rates expected when using a particular set of genetic markers for pairwise relationship inference. It also
provides functions implemented in C++ for computing likelihood ratios for different 
relationships between all pairs of individuals in a data set.  

CKMRsim can be installed with the 'remotes' packages`  It requires that development tools be installed on your
system to allow compilation of C++ via RCpp.
```{r, eval=FALSE}
remotes::install_github("eriqande/CKMRsim", build_vignettes = TRUE)
```

Once that is done (and note that building the vignettes can take a while), you can read the vignettes
with:
```{r, eval=FALSE}
vignette("CKMRsim-example-1")

vignette("CKMRsim-example-2-microsatellites")

vignette("CKMRsim-simulating-linked-markers")

vignette("CKMRsim-writing-geno-error-funcs")

vignette("CKMRsim-pairwise-relationships")

```

An easier way to get to read these vignetttes is to find them in the "articles"
tab at CKMRsim's pkgdown page at: https://eriqande.github.io/CKMRsim/

Simulations with physically linked markers are now performed internally using
CKMRsim's Rcpp gene-dropper by default, so Mendel is no longer required for
routine linked-marker simulations. The legacy Mendel backend remains available
with `useMendel = TRUE` and `install_mendel()`, mainly for reproducing older
analyses or backend comparisons, but it is no longer the recommended path.
