---
title: "CKMRsim"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This package implements Monte Carlo methods
(including efficient importance sampling approaches) for assessing the false positive and false
negative rates expected when using a particular set of genetic markers for pairwise relationship inference. It also
provides functions implemented in C++ for computing likelihood ratios for different 
relationships between all pairs of individuals in a data set.  

CKMRsim can be installed with devtools.  It requires that development tools be installed on your
system to allow compilation of C++ via RCpp.
```{r, eval=FALSE}
devtools::install_github("eriqande/CKMRsim", build_vignettes = TRUE)
```

After this, the various vignettes can be accessed and read like this:
```{r, eval=FALSE}
vignette("CKMRsim-example-1")

vignette("CKMRsim-example-2-microsatellites")

vignette("CKMRsim-simulating-linked-markers")

vignette("CKMRsim-writing-geno-error-funcs")
```

In order to pursue simulations in the face of physical linkage, you must download
and install the external dependency, Mendel version 16.  This was formerly
available at [http://software.genetics.ucla.edu/mendel](http://software.genetics.ucla.edu/mendel), but it seems that the site is no longer maintained or connected.

Instead, you can now install it the `install_mendel()` function.

