---
title: "CKMRsim-pairwise-relationships"
author: "Eric C. Anderson"
date: "`r Sys.Date()`"
resource_files:
  - images/AN.png
  - images/DFC.png
  - images/DHFC.png
  - images/FC.png
  - images/FS.png
  - images/GP.png
  - images/HAN.png
  - images/HFC.png
  - images/HS.png
  - images/HSC.png
  - images/PO.png
  - images/SC.png
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CKMRsim-pairwise-relationships}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

CKMRsim is all about inference of pairwise relationships from genetic data. This vignette documents
the named pairwise relationships that come "pre-loaded" with CKMRsim.  By saying that they are
"pre-loaded" we mean that each relationship's IBD coefficients are included in the package's
`kappa` matrix and that the pedigree specifying the relationship (for simulating linked markers)
is included in the `pedigrees` list. Both `kappas` and `pedigrees` are included
as part of the built-in data for this package.

First, we shall print out a table of the identify coefficients for the different 
relationships.  
```{r, message=FALSE, warning=FALSE}
library(CKMRsim)
library(tibble)
library(dplyr)
library(stringr)

long_names <- c(
  MZ = "Monozygotic (identical) twins",
  PO = "Parent-offspring",
  FS = "Full siblings",
  HS = "Half siblings",
  GP = "Grandparent-grandchild",
  AN = "Aunt-niece",
  DFC = "Double first cousins",
  HAN = "Half aunt-niece",
  FC = "First cousins",
  HFC = "Half first cousins",
  DHFC = "Double half first cousins",
  SC = "Second cousins",
  HSC = "Half second cousins",
  U = "Unrelated"
)

rtable <- as_tibble(kappas) %>%
  mutate(
    abbreviation = rownames(kappas),
    relationship = long_names
    ) %>%
  select(relationship, abbreviation, everything())

rtable
```

And now we will show a series pedigrees that depict the relationships, 
(except for monozygous twins and unrelated, which don't yield interesting
pedigrees) alongside the IBD
coefficient values. In these pedigrees,
the nodes corresponding to members of the pair are shaded (gray) and are labeled 1 and 2.

The members of each pair are labelled 1 and 2 in these pedigree definitions.
CKMRsim now uses these same pedigrees with its internal Rcpp linked-marker
simulator by default. The older Mendel-based backend is still available for
legacy comparisons, and these pedigrees retain the non-founder convention that
made the Mendel output parsing straightforward.


```{r, results='asis', echo=FALSE}
for (i in 2:(nrow(rtable) - 1)) {
  tit <- str_c("\n\n## ", rtable$relationship[i], "\n")
  kap <- str_c(
    "\n$\\kappa_0 = ", rtable$kappa0[i], 
    " ~~~~~~~\\kappa_1 = ", rtable$kappa1[i], 
    " ~~~~~~~\\kappa_2 = ", rtable$kappa2[i], 
    "$\n")
  fig <- str_c("\n![", rtable$abbreviation[i], ":  ", rtable$relationship[i], "](images/", rtable$abbreviation[i],  ".png)\n\n")
  cat(tit, kap, fig)
}
```
