---
title: "CKMRsim-pairwise-relationships"
author: "Eric C. Anderson"
date: "`r Sys.Date()`"
resource_files:
  - images/AN.png
  - images/DFC.png
  - images/DHFC.png
  - images/FC.png
  - images/FS.png
  - images/GP.png
  - images/HAN.png
  - images/HFC.png
  - images/HS.png
  - images/HSC.png
  - images/PO.png
  - images/SC.png
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CKMRsim-pairwise-relationships}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

CKMRsim is all about inference of pairwise relationships from genetic data. This vignette documents
the named pairwise relationships that come "pre-loaded" with CKMRsim.  By saying that they are
"pre-loaded" we mean that each relationship's IBD coefficients are included in the package's
`kappa` matrix and that the pedigree specifying the relationship (for simulating linked markers)
is included in the `pedigrees` list. Both `kappas` and `pedigrees` are included
as part of the built-in data for this package.

First, we shall print out a table of the identify coefficients for the different 
relationships.  
```{r, message=FALSE, warning=FALSE}
library(CKMRsim)
library(tibble)
library(dplyr)
library(stringr)

long_names <- c(
  MZ = "Monozygotic (identical) twins",
  PO = "Parent-offspring",
  FS = "Full siblings",
  HS = "Half siblings",
  GP = "Grandparent-grandchild",
  AN = "Aunt-niece",
  DFC = "Double first cousins",
  HAN = "Half aunt-niece",
  FC = "First cousins",
  HFC = "Half first cousins",
  DHFC = "Double half first cousins",
  SC = "Second cousins",
  HSC = "Half second cousins",
  U = "Unrelated"
)

rtable <- as_tibble(kappas) %>%
  mutate(
    abbreviation = rownames(kappas),
    relationship = long_names
    ) %>%
  select(relationship, abbreviation, everything())

rtable
```

And now we will show a series pedigrees that depict the relationships, 
(except for monozygous twins and unrelated, which don't yield interesting
pedigrees) alongside the IBD
coefficient values. In these pedigrees,
the nodes corresponding to members of the pair are shaded (gray) and are labeled 1 and 2.

Due to the way output files from Mendel are parsed.  It is necessary that the members of the
pair not be founders in the pedigree.  This is why the pedigrees for GP and PO have
founders above the oldest members of the pair.


```{r, results='asis', echo=FALSE}
for (i in 2:(nrow(rtable) - 1)) {
  tit <- str_c("\n\n## ", rtable$relationship[i], "\n")
  kap <- str_c(
    "\n$\\kappa_0 = ", rtable$kappa0[i], 
    " ~~~~~~~\\kappa_1 = ", rtable$kappa1[i], 
    " ~~~~~~~\\kappa_2 = ", rtable$kappa2[i], 
    "$\n")
  fig <- str_c("\n![", rtable$abbreviation[i], ":  ", rtable$relationship[i], "](images/", rtable$abbreviation[i],  ".png)\n\n")
  cat(tit, kap, fig)
}
```

