---
title: "Simulating-linked-markers"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulating-linked-markers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# Keep the heavier linked-marker demo out of ordinary vignette builds. Linked
# marker simulation now runs internally via Rcpp by default; Mendel is only
# needed if the legacy backend is requested explicitly with useMendel = TRUE.
DoLinkedDemo <- FALSE
NoLinkedDemo <- !DoLinkedDemo
```

```{r setup}
library(CKMRsim)
library(tidyverse)
```


## Prerequisites

CKMRsim now simulates physically linked markers internally using its Rcpp
gene-dropper. No Mendel installation is required for the recommended/default
workflow.

The older Mendel-based backend remains available by setting `useMendel = TRUE`,
but it is retained mainly for reproducing older analyses and backend comparisons.
For new linked-marker simulations, the internal Rcpp backend is recommended.

## An Example

'CKMRsim' comes with an example data set of linked markers.  The data themselves are concocted---they are merely
the microhaplotypes that have been assigned some fictitious positions within a genome---but that is useful for
demonstration. The data are allele frequencies in the format required by 'CKMRsim'. Here are the first few
lines of them"
```{r}
linked_mhaps
```
See that each marker is given a position (in base pairs) along the chromosome upon which it resides.

By default, 'CKMRsim' assumes a simple recombination rate of 1 cM per megabase.  If you have an actual
genetic linkage map for you markers, then you can accommodate that by giving base-pair positions (`Pos`)
of the markers that correspond to the positions that would yield the observed recombination fractions
under a constant rate of 1 cM/Mb.

Here we will show the process of simulating full sibling pairs with and without the physical linkage
in `linked_mhaps`.  We will also show that physical linkage does not affect parent-offspring or
unrelated pairs.

## Simulating Pairs

Creating the CKMR object is just the same as it was in the "Example 1" vignette.
```{r}
ck_lmh <- create_ckmr(
  D = linked_mhaps,
  kappa_matrix = kappas[c("PO", "FS", "U"), ],
  ge_mod_assumed = ge_model_TGIE,
  ge_mod_true = ge_model_TGIE,
  ge_mod_assumed_pars_list = list(epsilon = 0.01),
  ge_mod_true_pars_list = list(epsilon = 0.01)
)
```

Simulating the markers under the assumption of no linkage is also the same:
```{r}
Qs_lmh_no_link <- simulate_Qij(ck_lmh,
                               calc_relats = c("PO", "FS", "U"),
                               sim_relats = c("PO", "FS", "U") 
                               )
```


In order to simulate the markers with physical linkage, you use the same
`simulate_Qij()` function, but you tell it to simulate with physical linkage
_and_ you also give it a list of pedigrees. By default, CKMRsim uses those
pedigrees with its internal Rcpp gene-dropper. The legacy Mendel backend can
still be requested with `useMendel = TRUE`, but that route is no longer
recommended for routine work.

`CKMRsim` comes with such a list. The two focal individuals of each pair are
individuals 1 and 2. If you want to add more pedigrees, like cousins, feel free.
If you need help, ping me at [eric.anderson@noaa.gov](mailto:eric.anderson@noaa.gov).

So, simulating with linkage looks like this:
```{r, eval=DoLinkedDemo}
# This chunk is skipped in ordinary vignette builds because it is stochastic and
# comparatively heavy, not because it requires Mendel.
Qs_lmh_with_link <- simulate_Qij(ck_lmh,
                                 calc_relats = c("PO", "FS", "U"),
                                 sim_relats = c("PO", "FS", "U"), 
                                 unlinked = FALSE,
                                 pedigree_list = pedigrees
)

```


Note that `CKMRsim` is wise enough to simply simulate markers as unlinked for the
PO, U, and MZ cases (because, when dealing with pairs with those relationships,
the presence of physical linkage makes no difference to the outcome).


Carrying on, it is instructive to look at the distributions of the FS/U log-likelihood
ratios.  They should differ between the cases with and without linkage, most notably the case
with linkage should produce values with greater variance.  Let us see if that
is the case. To do that, we extract the log-likelihood ratio values, put them together
in a single data frame, and then plot them.  Note that the `extract_logls` function
returns a column that says whether the simulation was done with or without linkage, so
it is straightforward to do this in two lines:
```{r, eval=DoLinkedDemo}
lnl_logls <- list(
  extract_logls(Qs_lmh_no_link, numer = c(FS = 1), denom = c(U = 1)),
  extract_logls(Qs_lmh_with_link, numer = c(FS = 1), denom = c(U = 1))
) %>%
  bind_rows()
```
```{r, eval=DoLinkedDemo, include=FALSE, echo=FALSE}
saveRDS(lnl_logls, file = "data/lnl_logls.rds")
```
```{r, eval=NoLinkedDemo, include=FALSE, echo=FALSE}
lnl_logls <- readRDS("data/lnl_logls.rds")
```

Now compare the cases when the truth is full-sibling:
```{r, fig.width = 7}
lnl_logls %>%
  filter(true_relat == "FS") %>%
  ggplot(aes(x = logl_ratio, fill = simtype)) +
  geom_histogram(bins = 70, 
                 position = position_identity(),
                 alpha = 0.5)
```


See, logl ratios have the same mean top and bottom.

```{r}
lnl_logls %>%
  filter(true_relat == "FS") %>%
  group_by(simtype) %>%
  summarise(mean_logl_rat = mean(logl_ratio),
            mean_numer = mean(numer_logl))
```
