---
title: "CKMRsim-writing-geno-error-functions"
author: "Eric C. Anderson"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{CKMRsim-writing-geno-error-functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

`CKMRsim` allows users to specify genotyping error models for locus $\ell$ in terms of a matrix $\mathbf{C}_\ell$.
If there are $G$ possible genotypes, then $\mathbf{C}_\ell$ is a $G \times G$
matrix where the element in row $s$ and column $t$ is the probability that an individual
truly carrying genotype $s$ is observed to have genotype $t$.  This model covers any possible
type of genotyping error in which errors are independent between different loci and are also
independent between different individuals.  That is a wide class of models.

`CKMRsim` comes with some "pre-packaged" genotyping error models, but users might want to
create their own that are tailored to the particular marker types that they work on. Currently,
the way to define a genotyping error model is to define a function that takes an input
`L`, which carries information about the locus, and any additional named parameters (like
genotyping error rates.)  At some point, I will reimplement most of the list structures in
`CKMRsim` as list-columns in tibbles, and then it will be much easier to define 
locus-specific, and even allele-specific, error rates, etc.  For now, it is a little tough
to do that cleanly and programmatically on a large scale...

In this vignette, I will show how I created genotyping error functions for two of the 
"pre-packaged" versions: True-Genotype-Independent Errors (TGIE), and a model tailored for
microsatellites that includes allele mis-calling and allele dropout.

## Basic Function Format

Every genotyping error model function should have a name starting with `ge_model_`.
For example we will name our TGIE model `ge_model_TGIE()`.  This naming convention is
not a requirement, but it helps to keep things organized.  

These `ge_model_` functions are required to have one argument named `L`.  This will be
where information about the locus gets passed into the function.  Currently, the information
available within `L` about the locus is contained in two list components:

* `L$freqs` : a named vector of allele frequencies.  The names of the vector are the
names of the corresponding alleles.  So, for example, if the names were microhaplotypes,
like "ACCAT", then that information would be available for use.  Alternatively, if the
names are microsatellite allele lengths, like "112", then those would also be
available for creating the genotyping error model. Note that the alleles are listed 
in this vector in the order in which CKMRsim represents them internally (for the C++ based
functions), so whenever you do anything with the alleles, you want to keep them in this
order!
* `L$geno_freqs` : a named vector of genotype frequencies ordered in the way that
`CKMRsim` has ordered the possible diploid genotypes that can be formed from the
different alleles.  The names are the names of the genotypes, which are merely the 
names of the two alleles in the genotype, separated by space-slash-space.
For example "112 / 116". (Note that, by default, CKMRsim names genotypes this way.
There should be no problem parsing such genotype names into the names of the constituent
alleles, so long as alleles are never named have a space.  That should be a problem.  If
you have spaces in your allele names, then please fix that!)
  

After the required `L` argument, the `ge_model_` function can have any other
arguments that control the behavior of the genotyping error.

The return value of the function must be the $G \times G$ matrix $\mathbf{C}_\ell$
described above. The rows of this matrix must sum to one, since they are
probabilities.  Note that $G$ is given by `length(L$geno_freqs)`. 


## The TGIE Function

True-genotype-independent error is a very simple genotyping error model that has been
in use for a long time in relationship inference.  The main goal of it is to ensure that
genotyping errors don't create zero-probability situations.  It is not intended to
model the actual genotyping error process, but it is mathematically tractable in a lot
of situations, and it is useful in cases where little is known about the genotyping error
process.  One such situation occurs when someone gives you a file of genotypes that are coded
up as integers, but those integers mean very little, i.e., they just denote separate alleles 
which could be microsatellite alleles (but not the actual lengths), or they could be
microhaplotype alleles (but not the actual variant sequences), or they could be SNPs (but
not the actual bases, etc).  

We will pretend for now that we are dealing with such integer-coded data and that we might have
multiple alleles at each locus.  For example, the components of `L` might look like this:
```r
> L$freqs
           1            2            3            4            5 
6.329010e-01 3.662781e-01 4.104416e-04 2.462650e-04 8.208833e-05 
           6 
8.208833e-05 
> L$geno_freqs
       1 / 1        1 / 2        1 / 3        1 / 4        1 / 5 
4.005637e-01 4.636356e-01 5.195378e-04 3.117227e-04 1.039076e-04 
       1 / 6        2 / 2        2 / 3        2 / 4        2 / 5 
1.039076e-04 1.341597e-01 3.006716e-04 1.804029e-04 6.013432e-05 
       2 / 6        3 / 3        3 / 4        3 / 5        3 / 6 
6.013432e-05 1.684623e-07 2.021548e-07 6.738493e-08 6.738493e-08 
       4 / 4        4 / 5        4 / 6        5 / 5        5 / 6 
6.064644e-08 4.043096e-08 4.043096e-08 6.738493e-09 1.347699e-08 
       6 / 6 
6.738493e-09 
```
This model is parameterized
by a genotyping error rate, $\epsilon$, and errors occur independently between 
individuals and loci.  With probability $1-\epsilon$ a genotyping error does not occur so that the observed
genotype is the true genotype. With probability $\epsilon$ the genotype is subject to genotying error, in
which case, the observed value is drawn from the remaining genotypes, proportionally to their frequencies.
Under such a model, using $y_\ell$ to denote the observed genotype and $x_\ell$ to denote the true genotype, we have
the row $s$ and column $t$ entry of $\mathbf{C}_\ell$ being the the conditional probability:
\[
P(y_\ell = g_{t} | x_\ell = g_{s}) = 
\left\{
\begin{array}{ll}
(1-\epsilon), &  s = t \\
\epsilon q(g_{t}) / (1 - g_s), & s \neq t.
\end{array}
\right.
\]
where $q(g_{s})$ is the frequency of genotype $s$.

Thus, to implement TGIE, all we need to do is write code to return the matrix
$\mathbf{C_\ell}$ given the input `L$geno_freqs` and another parameter `epsilon`.
That looks like this:
```r
#' implements a simple true-genotype-independent genotyping error model
#' @param L required locus specific information
#' @param epsilon the rate at which genotypes are incorrectly observed.
ge_model_TGIE <- function(L, epsilon = 0.01) {
  # first make a matrix G x G matrix of the genotype frequencies
  gmat <- matrix(rep(L$geno_freqs, length(L$geno_freqs)), byrow = TRUE, nrow = length(L$geno_freqs))

  # Now, we make the diagonals 1 - epsilon exactly and rescale
  # everything else to sum to epsilon.
  diag(gmat) <- 0
  gmat <- t(apply(gmat, 1, function(x) x * (epsilon/sum(x))))
  diag(gmat) <- 1 - epsilon

  # put the genotype names on for the row and column names since that can be
  # handy down the road.
  rownames(gmat) <- names(L$geno_freqs)
  colnames(gmat) <- names(L$geno_freqs)

  # return gmat, which is the matrix C.
  gmat
}

```
That is all there is to it.  And we can run it on a simple 2-allele example here:
```{r}
library(CKMRsim)
L <- list(
  freqs = c(A = 0.6, B = 0.4),
  geno_freqs = c(`A / A` = 0.36, `A / B` = 0.48, `B / B` = 0.16)
)

C <- ge_model_TGIE(L, epsilon = 0.02)

C
```

It is always good to check that your function returns a matrix whose rows
sum to one.
```{r}
rowSums(C)
```


## A Quick Microsatellite Model

Here we develop a microsatellite model that tries to capture a couple of features of microsatellites:

1. If you mis-call an allele, you are more likely to mis-call it as an allele that is close in length
to the true allele than one that is of very different length then the true allele.
2. Some alleles don't amplify reliably, and this seems to happen more often with longer
alleles (large-allele dropout), leaving the remaining allele to appear to be in homozygous form.

Anyway, we note here that we have both allelic mis-calls and dropout errors.  CKMRsim includes a function
called `combine_miscalls_and_dropouts()` that makes it fairly easy to implement the combination of these
two types of errors, using a simple model (see the paper Anderson, in prep, for details.)  In order to use
this function one needs to specify two different sets of probabilities:

1. Mis-call probabilities: a matrix $\mathbf{W}$ that is $A \times A$ where $A$ is the number of alleles at the locus.
The element in row $i$ and column $j$ is the probability that a true allele of type $i$ is called as type $j$.
2. Dropout probabilities: a vector $D$ of length $A$ giving the probability that each different allele will
drop out and not be observed/called (yielding a genotype that is homozygous for the remaining allele.)

It is important to note that each of the two copies of a gene at a locus has a chance to drop out or be
miscalled.  Thus, if you have a true genotype of $CD$ and the $C$ allele has a dropout rate of 0.01 and 
$D$ a dropout rate of 0.02, then, a dropout will occur at the locus with probability $0.01 + 0.02 = 0.03$.

To develop this model we will want to refer to an example locus.  We use one of
the ones from the Ruzzante et al. data set in one of the brook trout populations.
This is here:
```{r}
example_L_microsat
```
That shows what the `L` argument to our genotyping error model might 
look like.  

When we implement this, here is how we get the features we describe above.
First, to get the matrix $\mathbf{W}$, taking account of proximity of 
different alleles (in terms of length) we do this:

* Allow the user to specify a basic underlying probability that an allele
is mis-called.
* Identify the step size between alleles as the minimum different between any
two allele lengths.
* If a mis-call error occurs, model the probability that an allele of length $X$
is mistakenly called an allele of length $Y$ is proportional to the geometric 
probability, with parameter $p$ (note that the user must set the parameter $p$).
\[
P(X~\mbox{is called as }Y) \propto p(1 - p)^{|Y-X| - 1}
\]
where $|Y-X|$ is the unsigned difference between $Y$ and $X$ in units of the 
step size.  Note that if all allele lengths are a multiple of the step size, then
this would be an equality, not just a proportion, but, in order to allow for
intervals between alleles that might not be multiples of the step size, we will
just allow for $|Y-X|$ to be a real number, and we will normalize over all possible
alleles.  An interesting consequence of this normalization is that genotyping error probabilities
may not be symmetrical: if allele 112 is adjacent to 116 and 120, but there is
no 108 or 104, then 112 is more likely to be miscalled as 116 than 116 is to be
miscalled as 112, because there is also a good chance that it will be miscalled as
120 (whereas 112 doesn't have many likely miscall states below it.)


Then, to get some sort of "large-allele dropout" effect, we will do this:

* Have the user set a "base" drop out rate, $d$, which is the rate at which a 
"typical" allele would drop out.
* From the observed frequency distribution of alleles, compute the mean and the
standard deviation of the allele-length distribution.
* For each allele, compute a $z$-score that describes where it fits in that
allele-length distribution.
* For each allele with $z>0$ compute the dropout rate as $dsz$ where $d$
is the base dropout rate, $s$ is a scaling factor that controls how much
dropout rate increases with allele length, and $z$ is the allele's 
$z$-score.

Here is a function that accomplishes that.  This is just the function
definition from the package.
```{r, eval=FALSE}
#' a simple "length-aware" genotyping error model for microsatellites
#'
#' In this genotyping error model, mis-calls are more likely to nearby
#' allele lengths than distant ones, and larger alleles are more likely
#' to drop out than small or average-sized ones.
#'
#' @param L an element of the list created by \code{\link{long_markers_to_X_l_list}}. Such
#' an element basically holds the information at a single locus.  The idea here
#' is that every ge_mod_* function takes in an object like L, and then can
#' use any piece of information in it about alleles or genotypes to
#' configure a genotyping error model. In the case of microsatellites,
#' the names of the alleles must be the allele lengths, like "64" or "112".
#' @param miscall_rate the rate at which alleles are miscalled.
#' @param miscall_decay the parameter (or a geometric distribution) that determines
#' how quickly the probability of the miscall being to a certain allele length
#' decreases as you get further and further away from the true allele length.
#' The larger this values, the more quickly the probability decays.  For every
#' microsatellite allele step size further from the true allele size, the probability
#' increases by a factor of 1 - miscall_decay.  Thus, if miscall_decay = 1
#' then all errors are a step-size of one away, and if it is 0, then
#' the miscall is equally likely to occur to any other allele length.
#' @param dropout_rate the base rate at which alleles of "typical" size dropout.
#' @param dropout_scale_factor the rate of drop out increases for alleles
#' larger than the (frequency-weighted) average size of allele by a factor of
#' dropout_scale_factor times the z-score of the allele length.
#' @export
#'
ge_model_microsat1 <- function(L,
                               miscall_rate = 0.005,
                               miscall_decay = 0.8,
                               dropout_rate = 0.01,
                               dropout_scale_factor = 0.25
) {

  # first, compute the step-size as the minimum distance between any two alleles
  lengths <- as.integer(names(L$freqs))
  diffs <- outer(lengths, lengths, function(x,y) abs(x - y))
  diag(diffs) <- NA
  min_diff = min(diffs, na.rm = TRUE)

  step_size = min_diff

  # now, we compute the decay factors between all pairs of alleles,
  # and then we multiply that by the base-miscall rate to get the
  # matrix W of allele-to-allele miscall rates.
  steps <- diffs/step_size

  preW <- miscall_decay * (1 - miscall_decay) ^ (steps - 1)

  # and those need to be normalized so that the off-diagonals sum to miscall_rate
  W <- t(apply(preW, 1, function(x) x * miscall_rate / sum(x, na.rm = TRUE)))
  # and we put 1 - miscall_rate on the diagonal
  diag(W) <- 1 - miscall_rate
  # and we MUST put the allele lengths on there as the names
  rownames(W) <- names(L$freqs)
  colnames(W) <- names(L$freqs)



  # Now, let's make the vector D of dropout rates.
  # first get the mean allele length
  mean_len <- sum(lengths * L$freqs)

  # and then we get the variance and the standard deviation.
  # Note that we have the freqs of each lengths, but not the sample
  # sizes, so we can't do the n-1 sort of correction for variance.
  var_len <- sum( L$freqs * (lengths - mean_len) ^ 2)
  sd_len <- sqrt(var_len)

  # now compute the z-scores
  zscores <- (lengths - mean_len) / sd_len

  # now we can compute the vector D of dropout probabilities
  D <- rep(dropout_rate, length.out = length(L$freqs))
  D[zscores > 0] <- dropout_rate * (1 + zscores[zscores > 0] * dropout_scale_factor)
  names(D) <- names(L$freqs)

  # And now we create the matrix C by pushing all these into the function
  # that combines allelic mis-calls and dropouts. Note that by setting Ws
  # to W here we assume that the occurrence of a dropout does not affect the
  # rate at which mis-calls occur.
  ret <- combine_miscalls_and_dropouts(geno_names = names(L$geno_freqs),
                                D = D,
                                W = W,
                                Ws = W)
  
  ret

}

```

We can run that function with the default values to see what the results look like.
The matrix is quite large (there are many genotypic states) but so we just look at the
first 4 rows and columns.
```{r}
default <- ge_model_microsat1(example_L_microsat)
default[1:4, 1:4]
```
Note that, because of the way dropout errors occur, you are more likely to
encounter a genotyping error when the true genotype is a heterozygote.

It will be worth the user's time to play around with different settings to see the
effect of them.


## A Genotyping Error Model for Microhaplotypes

Microhaplotypes are clusters of a handful of SNPs that are all close enough to one another that
they can be assayed by amplifying the segment of DNA they are on and then sequencing it on a 
next generation sequencing machine.  Since such machines are given the sequence of a single piece
of DNA (typically about 100 to 300 bp long), the alleles that occur together on such a read are certain
to have been together on the same haplotype.  These microhaplotypes are thus multiallelic markers. 
The main type of genotyping error that occurs with them is a dropout error, in which an individual
carries one or two copies of the gene that do not amplify well, and thus will not yield enough reads to
detect that allele.  On top of that, sequencing errors might lead one to read an incorrect sequence
for one of the alleles.  In this case, it is more likely that the erroneous haplotype is "close" to the
true one (i.e., differs at few SNPs), rather than far from it (different at many SNPs from the true
haplotype.)

The alleles of a microhaplotype can be named by the allelic type of the SNPs within it.  For example,
at a microsatellite composed of 5 SNPs, one allele might be "ACCGT".  Here we show a simple
error model for genotypes involving such microhaplotype alleles.  It has the following features:

1. The locus has a single allele mis-call rate, call it `miscall_rate`.  In some circumstances,
we might want loci with many SNPs to have a higher miscall rate.  This is achieved by multiplying
the miscall rate by the number of SNPs.  
2. If a miscall occurs, with allele $X$ being mis-called as allele $Y$, then the probability of miscalling
$Y$ when the truth is $X$ is inversely proportional to the number of SNP alleles at which $X$ and $Y$ 
differ.
3. There is a locus-specific allelic dropout rate $d$.  In a future release we will make it
less cumbersome to allow allele-specific dropout rates.
4. The allelic mis-calls and dropouts are combined using the function we have seen 
before, `combine_miscalls_and_dropouts()`.

We have, in the package, an example of what a microhaplotype locus might look like:
```{r}
example_L_microhap
```

So, we will need a function to compute distances between the microhap alleles,
and then we will need to use that to compute a matrix $\mathbf{W}$, like we did
for microsatellites, and then we will combine that with dropout rates.  In this function
those dropout rates are made the same for every allele, but that could be
changed.

Here is what such a function looks like:
```{r, eval=FALSE}

```

