---
title: "CKMRsim-example-1"
author: "Eric C. Anderson"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{CKMRsim-example-1}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

In this example, we will work through a relatively complete analysis of data from 
kelp rockfish.  These data have been used in two publications to date
[@baetscherDispersalNearshoreMarine2019; @baetscherMicrohaplotypesProvideIncreased2018].
The genotype data are included in the package as a gzip compressed csv file.  The
contents are identical to what can be downloaded from Dryad at:
[http://datadryad.org/bitstream/handle/10255/dryad.205630/kelp_genos_used_two_col.csv?sequence=1](http://datadryad.org/bitstream/handle/10255/dryad.205630/kelp_genos_used_two_col.csv?sequence=1).

Before we get started let's load the packages we need.  We'll be using a lot of
functions from the tidyverse, so let's do it.
```{r setup}
library(tidyverse)
library(CKMRsim)
```

## Reading in the Genotype Data set

We first must read in the genotypes of the individuals. This file holds genotypes of all the adults
and all of the juveniles.  The first part of our analysis involves simulations using allele frequencies.
Since we don't expect large full sibling groups in our data, it is reasonable to use both adults _and_ 
juveniles, together, to estimate allele frequencies in this kelp rockfish population. 

It is
also worth noting that, in general, it is important to include most/all of the individuals
you will be using in your CKMR study when estimating allele frequencies.  This is because
all alleles that you might encounter when doing kin-finding must have been accounted for and
must have allele frequencies computed for them.  Especially when using multiallelic markers like
microhaplotypes or microsatellites, it is important to make sure that every allele that you
might encounter during kin-finding is accounted for during the allele frequency and power estimation
phases of the analysis.

The file that holds the genotypes comes with 
the `CKMRsim` package at:
```{r}
geno_path <- system.file("extdata/kelp_genos_used_two_col.csv.gz", package = "CKMRsim")
```
When we read that in, we get a lot of warnings (not shown) that column names have been deduplicated. This
is because each locus in the file occupies two columns and the names of those columns are the 
same (i.e., the name of the locus.)
```{r, warning=FALSE}
genos1 <- read_csv(geno_path)
```

Let's look at the first bit of the resulting tibble:
```{r}
genos1[1:10, 1:10]
```

That is not a super informative print out.  But it does show us that the `NMFS_DNA_ID` column is
the identifier for the indvidual fish, and that the genetic data start in column 2.  Also, the
even columns have the locus name and the odd columns (after the first one) are named with the
locus names plus a `_1` appended to them.  The alleles of each genotype are recorded as numbers.

## Estimating Power for Relationship Inference

Estimating the power that this set of markers is expected to give us for
pairwise relationship inference involves doing Monte Carlo simulations from
the allele frequencies in the population.  So, we are going to first have
to estimate the allele frequencies from our genotype data set. And, we
have to present those allele frequencies in a specific format that
`CKMRsim` is expecting.

### Allele Frequency Input Format

`CKMRsim` is designed to be able to work with information about markers that
are in known locations on different chromosomes within the genome.  Note that
genome coordinates of markers _are not required_ but the input format has a place
to record the chromome and position of each marker.  If these are not known,
we fill them in with dummy values, as will be shown.  If simulations are done
without physical linkage, then the values used have no effect.

The package data `long_markers` shows an example of the allele frequency
format that `CKMRsim` expects.  It looks like this:
```{r}
long_markers
```

The main columns are as follows:

* `Chrom` should be a character or integer denoting which chromosome the marker
is on.  (For example `1` or `X` or `Omy12`)
* `Locus` gives a character vector with the names of the markers/loci.  Please don't use spaces
in the marker names!
* `Pos` gives the genome coordinates of the marker.  This needs to be a number (double or integer)
* `Allele` gives the name of each allele at each locus.  This _must_ be a character vector.  So,
in our example, we will coerce the numbers we have for each locus into a character vector. Please
do not use spaces in allele names.
* `Freq` is the frequency of each allele in the population.  These should sum to 1.0 over each
locus.

The remaining columns, `LocIdx` and `AlleIdx` are integer indices that get assigned to each
locus and to each allele within each index.  These columns get filled by using the 
`reindex_markers()` function.

### Computing Allele Frequencies from Genotype Data

The strategy for doing this in the tidyverse is to make a long data frame with columns
`NMFS_DNA_ID`, `Locus`, `gene_copy` and `Allele`, and then summarising as appropriate.  Notice that, 
since these fish are diploids, we will let `gene_copy` be `a` or `b` to denote whether we are
talking about the allelic type of the first or second gene copy of the locus in a diploid individual.
To do this, we will rename the column headers to look like `Locus.gene_copy`, then do some gathering
and separating:
```{r}
# make note of the current order of loci in the data set
nc <- ncol(genos1)
loci <- names(genos1)[seq(2, nc, by = 2)]

#  reset the locus names
names(genos1)[seq(3, nc, by = 2)] <- str_c(names(genos1)[seq(2, nc, by = 2)], "b", sep = ".")
names(genos1)[seq(2, nc, by = 2)] <- str_c(names(genos1)[seq(2, nc, by = 2)], "a", sep = ".")

# then make some long format genotypes
long_genos <- genos1 %>% 
  gather(key = "loc", value = "Allele", -NMFS_DNA_ID) %>%
  separate(loc, into = c("Locus", "gene_copy"), sep = "\\.") %>%
  mutate(Allele = as.character(Allele))
```
Now we can compute the allele frequencies, and while we are at it, we will say that 
these markers are all on chromosome `Unk` in positions given by their order in the
original data set.  And we add columns for AlleIdx and LocIdx, as required to
pass this tibble into `reindex_markers()`, but we can leave each entry as NA.
```{r}
alle_freqs <- long_genos %>%
  count(Locus, Allele) %>%
  group_by(Locus) %>%
  mutate(Freq = n / sum(n),
         Chrom = "Unk",
         Pos = as.integer(factor(Locus, levels = loci))) %>%
  ungroup() %>%
  select(Chrom, Pos, Locus, Allele, Freq) %>%
  arrange(Pos, desc(Freq)) %>%
  mutate(AlleIdx = NA,
         LocIdx = NA)

# see what that looks like:
alle_freqs

```
 
Once that is done, we pass them through the `reindex_markers()` function to yield
a data frame of allele frequencies that `CKRMsim` can use to do simulations.
```{r}
afreqs_ready <- reindex_markers(alle_freqs)
```

We now have what we need to do a simulation analysis of power for
pairwise relationship inference from these markers.

Note that we will come back to `long_genos` later to actually do the
pairwise relationship inference.

### Create a CKMR object (`create_ckmr()`)

The first step in the simulation process is to create an object of class `ckmr` which
holds a large number of matrices useful for simulating genotypes (and thus for estimating
power for relationship inference).  The main function that
we use for this is `create_ckmr()`.  This function takes, as input, the allele frequencies
formatted as we have above, and a few other settings having to do with the relationships 
that the user is interested in simulation and the genotyping
error models that the user want to consider, and then it creates all the necessary elements
for doing the simulations.

Specifically, the arguments you need to pass to `create_ckmr()` are as follows:

* `D` : the tibble of allele frequencies that has been run through `reindex_markers()`.
* `kappa_matrix` : A matrix that describes the pairwise relationships that you will be wanting
to jointly simulate genotypes and likelihoods for.  Each row is named by what you want to 
call the relationship and there are three columns which give, respectively, the probability
that a pair with such a relationship share 0, 1, or 2 genes _identical-by-descent_.  The CKMRsim
package comes with a matrix called `kappas` that has this information for 12 relationships:
    - **MZ** : monozgotic twins (or "self").  This can be used to figure out how much power you
    have for identifying the same individual, sampled twice.
    - **PO** : parent-offspring.
    - **FS** : full siblings.
    - **HS** : half siblings.
    - **GP** : grandparent - grandoffspring.
    - **AN** : aunt-neice (same as uncle-nephew or any such avuncular relationship)
    - **DFC** : double first cousins.
    - **FC** : first cousins.
    - **HC** : half cousins.
    - **U** : unrelated.
The `kappas` matrix looks like this:
    ```{r}
    kappas
    ```
* `ge_mod_assumed` : a function that describes the genotyping error model that will be applied to
the simulated data when computing the likelihoods of the genotypes. In this case we will use
`ge_model_TGIE` which is appropriate for integer-coded data. More information about this is in the
vignette about writing genotyping error functions: [the-tgie-function](CKMRsim-writing-geno-error-funcs.html#the-tgie-function).
* `ge_mod_true` : a function that describes the genotyping error model that will actually be used to
simulate the genotype data.  Being able to separately specify these two models (assumed and true)
allows the user to investigate the effects of misspecification of the genotyping error model.
* `ge_mod_assumed_pars_list` a list of named parameters for the assumed genotyping error model (or,
leave blank to use the defaults).
* `ge_mod_true_pars_list` a list of named parameters for the true genotyping error model (or,
leave blank to use the defaults).

For the present example we are going to start with the assumption that the true and the assumed
genotyping error models are the same: true-genotype-independent model with error rate (epsilon)
equal to 1 in 100.  We will be concerned with how well we can distinguish Parent-offspring pairs
from unrelated, and full-sibling pairs from unrelated, and we also want to see if half-siblings might
sometimes look like full siblings.  So, we will create this ckmr object with the following relationships:
PO, FS, HS, U. We do all that like this:
```{r}
ex1_ckmr <- create_ckmr(
  D = alle_freqs,
  kappa_matrix = kappas[c("PO", "FS", "HS", "U"), ],
  ge_mod_assumed = ge_model_TGIE,
  ge_mod_true = ge_model_TGIE,
  ge_mod_assumed_pars_list = list(epsilon = 0.01),
  ge_mod_true_pars_list = list(epsilon = 0.01)
)
```
Now, when we print this object we get a little summary of what is in it:
```{r}
ex1_ckmr
```
This is an S3 object of class `ckmr`.

### Simulate genotype pairs and calculate their log-probabilities (`simulate_Qij()`)

Once we have created our `ckmr` object, we can use it to efficiently simulate the
multilocus genotypes of pairs of individuals (of different relationships).  At the same
time, we can calculate the log-probability of those genotypes of simulated pairs of individuals
under the assumption that they are of one relationship type or another.  Those log genotype
probabilities are the main ingredient for computing the log-likelihood ratios used for
inferring relationships.  

The main function that we use in this step is `simulate_Qij()`.  This function simulates
genotypes from different relationships, and then for each simulated genotype pair it also
calculates the log probability of the pair of genotypes _conditional on the pair being_
of one or several relationships.
This function has four main inputs that we discuss here (the others allow the addition of
missing data and physical linkage, and are discussed in other vignettes).

* `C` : the `ckmr` object to use for simulation and probabilty calculation.
* `sim_relats` : the set of true relationships you want to simulate genotypes from.  
* `calc_relats` : the set of assumed relationships you wish to compute genotype probabilities
for, from the simulated data.
* `reps` : for each relationship in `sim_relats`, the number of genotype pairs to simulate. This
is, by default 10,000.

The arguments `sim_relats` and `calc_relats` can use some extra discussion here.  Keep in mind that
we are going to use the output of this function to learn about the distribution of likelihood
ratios.  For example, if we are trying to identify parent-offspring pairs, we will typically
use the ratio of he probability of a pair's genotypes given that they are parent-offspring, 
divided by the probability of the pair's genotypes given that they are unrelated.  We can denote
such a likelihood ratio by PO/U.  If we are looking for full sibling pairs, we would, rather,
typically use the likelihood ratio FS/U.  If, on the other we hand we wanted to know how well
full-siblings might be resolved from half-siblings, we could use FS/HS.  Whatever the
target of one's question, in order for CKMRsim to calculate such likelihood ratios from the
simulated data, it is necessary to compute the probabilities that are used in the ratios---that 
is what the `calc_relats` argument is for.  You should set it to
a vector of the relationships that occur in the
numerator or the denominator of the likelihood ratios that you might want to investigate.  For example,
if we want to investigate the distribution of the ratios PO/U, FS/U, and FS/HS, then
`calc_relats` should be `c("PO", "FS", "HS", "U")`.

`sim_relats` refers to the set of relationships that pairs of individuals are simulated
from.  You will typically be interested in the distribution of a likelihood ratio
_when the truth is relationship X_.  For example, if doing parentage inference, then to compute
false positive and false negative rates, you will need to know the distribution of PO/U 
when pairs are truly unrelated (U) and when they are truly parent-offspring, (PO).  But, you
might also be interested to know what the distribution of PO/U looks like when the pairs
are actually full-siblings.  If you were interested in all of the above, then you would set
`sim_relats` equal to `c("PO", "FS", "U")`.

In this vignette, for our current example, we will assume that we want to know
whether our set of markers has sufficient power for identifying PO pairs and FS
pairs from U pairs by using the likelihood ratios PO/U and FS/U.  So, we will
clearly want `calc_relats = c("PO", "FS", "U")`.  But, let us also imagine that
we are curious whether individuals that are truly HS might be mistakenly identified
as FS; in such a case we will be curious about the distribution of FS/U when the 
truth is HS, so we will also need to simulate from HS.  Hence we
will use `sim_relats = c("PO", "FS", "HS", "U")

It is worth pointing out at this point that any relationship found in
`calc_relats` or `sim_relats` must have been included as a rowname in
the `kappa_matrix` argument to `create_ckmr()`!

Here we simulate all the genotype-pair probabilities we will need:
```{r}
ex1_Qs <- simulate_Qij(ex1_ckmr, 
                       calc_relats = c("PO", "FS", "U"),
                       sim_relats = c("PO", "FS", "HS", "U") )
```

If we print the resulting S3 object (of class `Qij`), we get a little information
about it:
```{r}
ex1_Qs
```

### Compute log-likelihood ratios to view them (`extract_logls()`)

This next step is not entirely necessary, but it helps to create
pictures to understand what we are doing here, and the `extract_logls()`
function lets us get simulated log-likelihood ratios out of the `Qij` 
object we created above and plot a histogram or a density plot
of the distributions.  This is helpful for developing
intuition about things and understanding what is going on.

The main function here is `extract_logls()`.  It simply takes an object of
class `Qij` and uses the information within it to compute log-likelihood ratios
from all the genotype probabilities that are stored within it.  There is one
small twist: the numerator and denominator of the likelihood ratios can be
mixtures of different relationship categories.  For example, instead of 
just computing PO/U, you might want to compute a log likelihood ratio that
reflects your belief that, of the genotyping pairs you are sampling, 
95% are unrelated and the other 5% are full siblings.  Then you might
want to base your inference on a log-likelihood ratio that looked like:
PO/(0.95 * U + 0.05 * FS).  

Often you will not want to do that, but, because `extract_logls()` is designed
to do that, you have explicitly tell it how much weight to give each
relationship category in the numerator and the denominator of the log likelihood
ratio. 

An example here should help.  Let's say we are interested in the distribution
of the log likelihood ratio PO/U.  Then we would do:
```{r}
PO_U_logls <- extract_logls(ex1_Qs,
                            numer = c(PO = 1),
                            denom = c(U = 1))
```




## Literature Cited
