---
title: "Developer Notes"
author: "Eric C. Anderson"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
library(CKMRsim)
```

These are some notes that Eric kept while putting this package together, mostly to remind
him of the flow of everything and to put little notes in describing how everything goes
together.  

## Marker definitions

Markers have to be defined in a data frame. Here is an example:
```{r, }
long_markers
```
The data frame has to be in this format.  It needs to have CHROM (which can be an int or string) 
Locus (chr) Pos (double or int), Allele need to be a character (even if it is an allele length).
The LocIdx and AlleleIdx fields give the order of things that is desired.  You can use
`reindex_markers` to sort things into the allele frequency order and re-index them. 

If you don't know the chrom or position you can just stick whatever you want in there.

## Genotyping error

There are two main types of genotyping error model, but so far I have only implemented the 
allele-based one:  `general_allele_based_geno_err_model`.  

Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
