---
title: "Compare Rcpp and Mendel linked-marker simulations"
output: html_document
---

```{r setup}
library(tidyverse)
library(CKMRsim)

set.seed(20260726)
```

This notebook compares CKMRsim's internal Rcpp linked-marker simulator to the
legacy Mendel-based simulator. It uses the public bearded seal SNP data from the
TWS CKMR workshop repository.

```{r data}
dir.create("data", showWarnings = FALSE)
beardo_path <- file.path("data", "beardo-genos.rds")

if(!file.exists(beardo_path)) {
  download.file(
    "https://raw.githubusercontent.com/eriqande/tws-ckmr-2022/main/data/beardo-genos.rds",
    destfile = beardo_path,
    mode = "wb"
  )
}

beardo_genos <- readr::read_rds(beardo_path)
dim(beardo_genos)
```

```{r allele-frequencies}
long_genos <- beardo_genos %>%
  rename(Indiv = Our_sample) %>%
  pivot_longer(
    cols = -Indiv,
    names_to = c("Locus", "gene_copy"),
    names_sep = "\\.",
    values_to = "Allele"
  )

locus_names <- unique(long_genos$Locus)

afreqs_ready <- long_genos %>%
  count(Locus, Allele) %>%
  group_by(Locus) %>%
  mutate(
    Freq = n / sum(n),
    Chrom = "Unk",
    Pos = as.integer(factor(Locus, levels = locus_names))
  ) %>%
  ungroup() %>%
  select(Chrom, Pos, Locus, Allele, Freq) %>%
  arrange(Pos, desc(Freq)) %>%
  mutate(AlleIdx = NA, LocIdx = NA) %>%
  filter(!is.na(Allele)) %>%
  reindex_markers()

afreqs_ready
```

Give the markers pseudo-genomic positions. The published lab notes used 16
autosomes for bearded seal and a roughly 2.4 Gb genome.

```{r pseudo-genome}
fake_genome <- geometric_chromo_lengths(
  n = 16,
  L = 2.4,
  sl = 0.25
)

afreqs_linked <- sprinkle_markers_into_genome(
  markers = afreqs_ready,
  genome = fake_genome$chrom_lengths
)

# Keep the demo reasonably quick. Increase this for a stronger comparison.
max_loci <- 600
keep_loci <- afreqs_linked %>%
  distinct(Chrom, Locus, Pos, LocIdx) %>%
  arrange(Chrom, Pos, LocIdx) %>%
  slice_head(n = max_loci) %>%
  pull(Locus)

afreqs_demo <- afreqs_linked %>%
  filter(Locus %in% keep_loci) %>%
  reindex_markers()

n_distinct(afreqs_demo$Locus)
```

Create a CKMR object and simulate FS pairs three ways:

1. unlinked markers,
2. linked markers with the internal Rcpp simulator,
3. linked markers with Mendel, when a Mendel binary is installed.

```{r simulate}
ck <- create_ckmr(
  D = afreqs_demo,
  kappa_matrix = kappas[c("PO", "FS", "HAN", "U"), ],
  ge_mod_assumed = ge_model_TGIE,
  ge_mod_true = ge_model_TGIE,
  ge_mod_assumed_pars_list = list(epsilon = 0.005),
  ge_mod_true_pars_list = list(epsilon = 0.005)
)

reps <- 1e04

profile_simulation <- function(label, expr) {
  expr <- substitute(expr)

  gc(reset = TRUE)
  timing <- system.time(
    value <- eval(expr, envir = parent.frame())
  )
  gc_after <- gc()

  mb_cols <- which(colnames(gc_after) == "(Mb)")
  max_r_heap_mb <- if(length(mb_cols) >= 3) {
    sum(gc_after[, mb_cols[3]])
  } else {
    NA_real_
  }

  list(
    value = value,
    profile = tibble(
      backend = label,
      user_sec = unname(timing[["user.self"]]),
      system_sec = unname(timing[["sys.self"]]),
      elapsed_sec = unname(timing[["elapsed"]]),
      max_r_heap_mb = max_r_heap_mb
    )
  )
}

unlinked_run <- profile_simulation(
  "unlinked",
  simulate_Qij(
    ck,
    sim_relats = "FS",
    calc_relats = c("FS", "U"),
    reps = reps,
    unlinked = TRUE
  )
)

rcpp_run <- profile_simulation(
  "rcpp",
  simulate_Qij(
    ck,
    sim_relats = "FS",
    calc_relats = c("FS", "U"),
    reps = reps,
    unlinked = FALSE,
    pedigree_list = pedigrees,
    useMendel = FALSE
  )
)

mendel_run <- profile_simulation(
  "mendel",
  simulate_Qij(
    ck,
    sim_relats = "FS",
    calc_relats = c("FS", "U"),
    reps = reps,
    unlinked = FALSE,
    pedigree_list = pedigrees,
    useMendel = TRUE
  )
)

Q_unlinked <- unlinked_run$value
Q_rcpp <- rcpp_run$value
Q_mendel <- mendel_run$value

simulation_profiles <- bind_rows(
  unlinked_run$profile,
  rcpp_run$profile,
  mendel_run$profile
)

simulation_profiles
```

```{r compare}
logls <- list(
  unlinked = extract_logls(Q_unlinked, numer = c(FS = 1), denom = c(U = 1)),
  rcpp = extract_logls(Q_rcpp, numer = c(FS = 1), denom = c(U = 1)),
  mendel = extract_logls(Q_mendel, numer = c(FS = 1), denom = c(U = 1))
)

logls <- bind_rows(logls, .id = "backend")

logls %>%
  group_by(backend) %>%
  summarise(
    n = n(),
    mean = mean(logl_ratio),
    sd = sd(logl_ratio),
    q025 = quantile(logl_ratio, 0.025),
    q500 = quantile(logl_ratio, 0.500),
    q975 = quantile(logl_ratio, 0.975),
    .groups = "drop"
  )
```

```{r plot, fig.width=8, fig.height=5}
ggplot(logls, aes(x = logl_ratio, fill = backend, colour = backend)) +
  geom_density(alpha = 0.25) +
  theme_bw()
```

For FS pairs, the linked simulations should have a noticeably larger variance
than the unlinked simulation. The Rcpp and Mendel linked distributions should be
close up to Monte Carlo error.

## HAN/U comparison

Now do the same comparison for half-avuncular (`HAN`) pairs, again using `U` as
the denominator.  This is the same unlinked/Rcpp/Mendel comparison as above, but
for a more distant relationship.

```{r simulate-han}
han_unlinked_run <- profile_simulation(
  "unlinked",
  simulate_Qij(
    ck,
    sim_relats = "HAN",
    calc_relats = c("HAN", "U"),
    reps = reps,
    unlinked = TRUE
  )
)

han_rcpp_run <- profile_simulation(
  "rcpp",
  simulate_Qij(
    ck,
    sim_relats = "HAN",
    calc_relats = c("HAN", "U"),
    reps = reps,
    unlinked = FALSE,
    pedigree_list = pedigrees,
    useMendel = FALSE
  )
)

han_mendel_run <- profile_simulation(
  "mendel",
  simulate_Qij(
    ck,
    sim_relats = "HAN",
    calc_relats = c("HAN", "U"),
    reps = reps,
    unlinked = FALSE,
    pedigree_list = pedigrees,
    useMendel = TRUE
  )
)

Q_han_unlinked <- han_unlinked_run$value
Q_han_rcpp <- han_rcpp_run$value
Q_han_mendel <- han_mendel_run$value

han_simulation_profiles <- bind_rows(
  han_unlinked_run$profile,
  han_rcpp_run$profile,
  han_mendel_run$profile
)

han_simulation_profiles
```

```{r compare-han}
han_logls <- list(
  unlinked = extract_logls(Q_han_unlinked, numer = c(HAN = 1), denom = c(U = 1)),
  rcpp = extract_logls(Q_han_rcpp, numer = c(HAN = 1), denom = c(U = 1)),
  mendel = extract_logls(Q_han_mendel, numer = c(HAN = 1), denom = c(U = 1))
)

han_logls <- bind_rows(han_logls, .id = "backend")

han_logls %>%
  group_by(backend) %>%
  summarise(
    n = n(),
    mean = mean(logl_ratio),
    sd = sd(logl_ratio),
    q025 = quantile(logl_ratio, 0.025),
    q500 = quantile(logl_ratio, 0.500),
    q975 = quantile(logl_ratio, 0.975),
    .groups = "drop"
  )
```

```{r plot-han, fig.width=8, fig.height=5}
ggplot(han_logls, aes(x = logl_ratio, fill = backend, colour = backend)) +
  geom_density(alpha = 0.25) +
  theme_bw()
```

For HAN pairs the signal is weaker than FS, but the linked Rcpp and Mendel
distributions should still agree closely up to Monte Carlo error.

## Pair-genotype index check

This check compares the marginal frequency of each simulated genotype-pair
index at each locus to the value expected from the corresponding `X_l` matrix.
It does not test linkage among loci, but it is a useful check on genotype-pair
indexing.

```{r pair-index-check}
expected_xl_table <- function(C, relat) {
  imap_dfr(C$loci, function(L, locus) {
    expected <- as.vector(L$X_l[[relat]])
    tibble(
      locus = locus,
      genotype_pair = seq_along(expected),
      expected = expected
    )
  })
}

sample_unlinked_true_pair_indices <- function(C, relat, num) {
  lapply(C$loci, function(L) {
    probs <- as.vector(L$X_l[[relat]])
    sample.int(
      n = length(probs),
      size = num,
      replace = TRUE,
      prob = probs
    )
  })
}

sample_linked_true_pair_indices <- function(D, ped, C, num, backend) {
  outgenos <- switch(
    backend,
    rcpp = CKMRsim:::sample_linked_true_genotype_pairs(
      df = D,
      ped = ped,
      num = num
    ),
    mendel = CKMRsim:::sample_linked_true_genotype_pairs_mendel(
      df = D,
      ped = ped,
      num = num
    ),
    stop("Unknown linked backend: ", backend)
  )

  locs <- names(C$loci)
  names(locs) <- locs

  lapply(locs, function(locus) {
    nG <- nrow(C$loci[[locus]]$X_l[[1]])
    nG * (outgenos$indiv2[, locus] - 1) + outgenos$indiv1[, locus]
  })
}

observed_pair_index_table <- function(gp, C, relat) {
  imap_dfr(gp, function(x, locus) {
    nstates <- length(as.vector(C$loci[[locus]]$X_l[[relat]]))
    tibble(
      locus = locus,
      genotype_pair = seq_len(nstates),
      simulated = tabulate(x, nbins = nstates) / length(x)
    )
  })
}

run_pair_index_check <- function(dataset, D, C, relat, num) {
  expected <- expected_xl_table(C, relat)

  map_dfr(c("unlinked", "rcpp", "mendel"), function(backend) {
    gp <- switch(
      backend,
      unlinked = sample_unlinked_true_pair_indices(
        C = C,
        relat = relat,
        num = num
      ),
      rcpp = sample_linked_true_pair_indices(
        D = D,
        ped = pedigrees[[relat]],
        C = C,
        num = num,
        backend = "rcpp"
      ),
      mendel = sample_linked_true_pair_indices(
        D = D,
        ped = pedigrees[[relat]],
        C = C,
        num = num,
        backend = "mendel"
      )
    )

    observed_pair_index_table(gp, C, relat) %>%
      left_join(expected, by = c("locus", "genotype_pair")) %>%
      mutate(
        dataset = dataset,
        relat = relat,
        backend = backend,
        .before = locus
      )
  })
}

data(linked_mhaps)

linked_mhaps_demo <- linked_mhaps %>%
  reindex_markers()

# We will check both FS and HAN genotype-pair probabilities with this data set.
ck_linked_mhaps <- create_ckmr(
  D = linked_mhaps_demo,
  kappa_matrix = kappas[c("FS", "HAN", "U"), ],
  ge_mod_assumed = ge_model_TGIE,
  ge_mod_true = ge_model_TGIE,
  ge_mod_assumed_pars_list = list(epsilon = 0.005),
  ge_mod_true_pars_list = list(epsilon = 0.005)
)


# do this with only 5000 reps
index_check_reps <- 5000

pair_index_check <- bind_rows(
  run_pair_index_check(
    dataset = "trimmed beardo SNPs",
    D = afreqs_demo,
    C = ck,
    relat = "FS",
    num = index_check_reps
  ),
  run_pair_index_check(
    dataset = "linked_mhaps",
    D = linked_mhaps_demo,
    C = ck_linked_mhaps,
    relat = "FS",
    num = index_check_reps
  ),
  run_pair_index_check(
    dataset = "trimmed beardo SNPs",
    D = afreqs_demo,
    C = ck,
    relat = "HAN",
    num = index_check_reps
  ),
  run_pair_index_check(
    dataset = "linked_mhaps",
    D = linked_mhaps_demo,
    C = ck_linked_mhaps,
    relat = "HAN",
    num = index_check_reps
  )
)

pair_index_check %>%
  group_by(dataset, relat, backend) %>%
  summarise(
    n = n(),
    mean_abs_diff = mean(abs(simulated - expected)),
    rmse = sqrt(mean((simulated - expected)^2)),
    max_abs_diff = max(abs(simulated - expected)),
    .groups = "drop"
  )
```

```{r pair-index-check-plot, fig.width=8, fig.height=6}
ggplot(pair_index_check, aes(x = expected, y = simulated)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey50") +
  geom_point(alpha = 0.25, size = 0.7) +
  facet_grid(dataset + relat ~ backend) +
  coord_equal() +
  theme_bw()
```

