---
title: "Compare Kappas and Pedigrees"
output: html_notebook
---

I just added a bunch of new pedigrees and a few new kappas to the package
and want to do a quick sanity check here.


```{r}
library(tidyverse)
library(CKMRsim)
```

First, make a linked markers CKMR object:
```{r}
# c("PO", "FS", "HS", "AN", "GP", "DFC", "FC", "HFC", "DHFC", "SC", "HSC", "U")
ckmr <- create_ckmr(
  D = linked_mhaps,
  kappa_matrix = kappas,
  ge_mod_assumed = ge_model_TGIE,
  ge_mod_true = ge_model_TGIE,
  ge_mod_assumed_pars_list = list(epsilon = 0.01),
  ge_mod_true_pars_list = list(epsilon = 0.01)
)
```


Now, we only want to sim from the focal relationship and calc for the
focal relationship and U. So:
```{r}
simcalc <- function(R = "FC", reps = 1e4) {
  Qs_no_link <- simulate_Qij(
    ckmr,
    calc_relats = c(R, "U"),
    sim_relats = c(R),
    reps = reps
  )
  Qs_link <- simulate_Qij(
    ckmr,
    calc_relats = c(R, "U"),
    sim_relats = c(R), 
    unlinked = FALSE,
    pedigree_list = pedigrees,
    reps = reps
  )
  vec <- 1;
  names(vec) <- R
  logls <- bind_rows(
    extract_logls(Qs_no_link, vec, c(U = 1)),
    extract_logls(Qs_link, vec, c(U = 1))
  )
  
  plot <- ggplot(logls, aes(x = logl_ratio, fill = simtype)) + 
    geom_histogram(bins = 100, position = "identity", alpha = 0.6) +
    ggtitle(R)
  
  list(logls = logls, plot = plot)
}


```

Now, let's try running that over all the different relationships:
```{r}
vals <- c("PO", "FS", "HS", "AN", "DFC", "FC", "HFC", "DHFC", "SC", "HSC")
names(vals) <- vals

plot_and_logls <- lapply(vals, function(R) {simcalc(R, reps = 5e4)})
```

Now we just want to make some plots. We want the pedigree in one panel and the logls
on the other.

```{r}
library(cowplot)

pednames <- c(
  AN = "aunt-niece", 
  DFC = "double-first-cousins", 
  DHFC = "double-half-first-cousins", 
  FC = "first-cousins", 
  HFC = "half-first-cousins", 
  HSC = "half-second-cousins", 
  SC = "second-cousins"
)

side_by_sides <- lapply(names(pednames), function(x) {

  pname <- pednames[[x]]
  pfile <- str_c("../pedigree-development/", pname, ".png")
  gplot <- plot_and_logls[[x]]$plot

  ped_fig <- ggdraw() + draw_image(pfile)
  plot_grid(ped_fig, gplot, align = "v", rel_heights  = 1, rel_widths = c(1, 2), nrow = 1)

})
names(side_by_sides) <- names(pednames)
```

Save these for putting in the talk:
```{r}
ggsave(side_by_sides$AN, filename = "../outputs/AN.png", height = 5, width = 9)
ggsave(side_by_sides$DFC, filename = "../outputs/DFC.png", height = 5, width = 9)
ggsave(side_by_sides$FC, filename = "../outputs/FC.png", height = 5, width = 9)
ggsave(side_by_sides$HFC, filename = "../outputs/HFC.png", height = 5, width = 9)
ggsave(side_by_sides$SC, filename = "../outputs/SC.png", height = 5, width = 9)
```

That will be good for my presentation.

Now, how about means and standard deviations:
```{r}
tmp <- lapply(plot_and_logls, function(x) x$logls) %>%
  bind_rows(.id = "relationship") %>%
 # filter(relationship != "PO") %>%
  group_by(relationship, simtype) %>%
  summarise(
    mean_LLR = mean(logl_ratio),
    sd = sd(logl_ratio),
    var = var(logl_ratio)
  ) %>%
  mutate(tmp = mean(mean_LLR)) %>%
  arrange(desc(tmp)) %>%
  select(-tmp, -sd)

tmp$mean_LLR = sprintf("%.2f", tmp$mean_LLR)
tmp$var = sprintf("%.2f", tmp$var) 

pivot_wider(
  tmp, 
  names_from = simtype,
  values_from = c(mean_LLR, var)
) %>% View()
```
