---
title: "Tracking Down the Genotyping Error Discrepancy With Physical Linkage"
output: html_notebook
---

Something is going awry, and it appears that genotyping errors are not
getting propagated into the physically linked data.

## Trying to Track it Down

For the first thing I am going to try is creating a CKMR object and then
using that to run one of the intermediate functions.
```{r}
library(CKMRsim)
library(tidyverse)

ck_lmh <- create_ckmr(
  D = linked_mhaps,
  kappa_matrix = kappas[c("PO", "FS", "U"), ],
  ge_mod_assumed = ge_model_TGIE,
  ge_mod_true = ge_model_TGIE,
  ge_mod_assumed_pars_list = list(epsilon = 0.000000005),
  ge_mod_true_pars_list = list(epsilon = 0.25)
)

# here I also make one with essentially no genotyping error
# because I want to see if that is what we are getting
ck_lmh2 <- create_ckmr(
  D = linked_mhaps,
  kappa_matrix = kappas[c("PO", "FS", "U"), ],
  ge_mod_assumed = ge_model_TGIE,
  ge_mod_true = ge_model_TGIE,
  ge_mod_assumed_pars_list = list(epsilon = 0.000000005),
  ge_mod_true_pars_list = list(epsilon = 0.0000000005) 
)
  
```

Once we have done that we can assign some variables to work through a function:
```{r}
r <- "FS"
YL <- ck_lmh$loci
reps <- 30000
names(reps) <- "FS"

df <- ck_lmh$orig_data

# these functions get run inside simulate_and_calc_Q()
# this is for the unlinked simulation
gpu <- lapply(YL, function(y) { # this is cycling over loci
        sample.int(n = length(y$Y_l_true[[r]]), size = reps[r], replace = TRUE, prob = y$Y_l_true[[r]])
      })


gpu2 <- lapply(ck_lmh2$loci, function(y) { # this is cycling over loci
        sample.int(n = length(y$Y_l_true[[r]]), size = reps[r], replace = TRUE, prob = y$Y_l_true[[r]])
      })

# and this is for the linked simulation
gpl <- sample_linked_genotype_pairs(df = ck_lmh$orig_data, ped = pedigrees[[r]], C = ck_lmh$loci, num = reps[r])
gpl2 <- sample_linked_genotype_pairs(df = ck_lmh2$orig_data, ped = pedigrees[[r]], C = ck_lmh2$loci, num = reps[r])

```

Now, we want to compare these things.  We can do some tidyr for that.
```{r}
gps <- bind_rows (
  linked = enframe(gpl),
  linked2 = enframe(gpl2),
  unlinked = enframe(gpu),
  unlinked2 = enframe(gpu2),
  .id = "type"
  ) %>%
  unnest(value)

# now, count up the number of occurrences
gcounts <- gps %>%
  count(type, name, value) %>%
  spread(key = type, value = n)

```

This is really weird.

## A new hypothesis

Perhaps `samp_from_mat()` is somehow using the same random number seeds, so that the
number of genotypes that are identical between the two individuals is abnormally high.

I can check on the fraction of of all simulated genes that have two gene copies identical
in state.  Those pairs of genotypes will be found along the diagonals.  So, let's count  those
diagonal genotypes up. First make a tibble of loci and the indices of the diaganals for Y_l:
```{r}
tmp <- lapply(ck_lmh$loci, function(x) {
  matrix(seq(1:length(x$Y_l$U)), nrow = nrow(x$Y_l$U), ncol = ncol(x$Y_l$U)) %>%
    diag()
})

diags <- tibble(
  name = names(tmp),
  diag = tmp
)
```

Then determine whether each value is on the diagonal or not:
```{r}
diag_suss <- gps %>%
  count(type, name, value) %>%
  left_join(diags, by = "name") %>%
  mutate(isDiag = map2_lgl(diag, value, function(x,y) y %in% x))

# and count up
diag_suss %>% 
  group_by(type) %>%
  summarise(fract_diag = sum(isDiag * n) / sum(n))

```

So, that is clearly too much identity by descent in the linked simulation.

So, let's just try samp_from_mat():
```{r}
# get a matrix to sample from.  Let it be totally random:
TR <- matrix(1/16, ncol = 16, nrow = 1e04)

g1 <- samp_from_mat(TR)
g2 <- samp_from_mat(TR)

mean(g1 == g2)
```
Nope.  That indicates that it isn't a case of failng to seed the RNG.

** WAIT! THE PROBLEM WAS WITH THE CALCULATION OF THE JOINT PROB (Y1,Y2) MATRIX
IN THE UNLINKED CASE.  HOLY CRAP! SO, THAT IS FIXED NOW AND THIS STUFF IS NO LONGER
TOTALLY GERMANE.  BUT WE WILL COMMIT IT FOR POSTERITY **

